Component response rate variation drives stability in large complex systems

### 2018-07-26 notebook/ms_for_review.Rmd
New files

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
    keep_tex: no
  html_document: default
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: science.csl
biblio-style: apalike
---

@@

@@

@@

@@

@@

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle centred at $\gamma = -1.95$ and one small circle centred at $\gamma = -0.05$. In a, the maximum real eigenvalue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` was stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-3.7, 0.3), ylim = c(-2, 2), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
A0x0 <- sqrt(200) * sd(A0vec) * cos(vl) + mean(diag(A0));
A0y0 <- sqrt(200) * sd(A0vec) * sin(vl);
text(x = -3.5, y = 2.25, labels = "a", cex = 2);
points(x = A0x0, y = A0y0, type = "l", lwd = 3, col = "grey");
points(A0_r, A0_i, pch = 4, cex = 0.7);

plot(A1_r, A1_i, xlim = c(-3.7, 0.3), ylim = c(-2, 2), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, 
     col = "black", yaxt = "n");

vl <- seq(from = 0, to = 2*pi, by = 0.001);
A0x1a <- sqrt(100) * sd(A1vec[fhalf]) * cos(vl) + mean(diag(A1)[1:100]);
A0y1a <- sqrt(100) * sd(A1vec[fhalf]) * sin(vl);
points(x = A0x1a, y = A0y1a, type = "l", lwd = 3, col = "grey");
A0x1b <- sqrt(100) * sd(A1vec[shalf]) * cos(vl) + mean(diag(A1)[101:200]);
A0y1b <- sqrt(100) * sd(A1vec[shalf]) * sin(vl);
points(x = A0x1b, y = A0y1b, type = "l", lwd = 3, col = "grey");

points(A1_r[1:100], A1_i[1:100],pch = 4, cex = 0.7);

@@

**Figure 2: Distributions of eigenvalues before (a) and after (b) introducing variation in component response rate ($\boldsymbol{\gamma}$) in complex systems.** Each panel show the same system where $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. **a.** Eigenvalues plotted in the absence of $Var(\gamma)$ where $E[\gamma] = 1$, versus **b.** eigenvalues plotted given $\gamma \sim \mathcal{U}(0, 2)$, which increases the variance of interaction strengths ($\sigma^{2}$) but also creates a cluster of eigenvalues toward the distribution's centre (-1, 0). Black elipses in both panels show the circle centred on the distribution in panel a. Proportions of $\Re(\lambda) < 0$ are `r sum(A0_r < 0) / length(A0_r)` and `r sum(A1_r < 0) / length(A1_r)` for a and b, respectively.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-6, 2.25), ylim = c(-5,4.25), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
text(x = -15.5, y = 19, labels = "a", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3);
plot(A1_r, A1_i, xlim = c(-6, 2.25), ylim = c(-5,4.25), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, col = "red",
     yaxt = "n");
text(x = -15.5, y = 19, labels = "b", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3, lty = "solid");
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

@@

**Figure 3: Stability of large complex systems with and without variation in component response rate($\boldsymbol{\gamma}$).** The $\ln$ number of systems that are stable across different system sizes ($S$) given $C = 1$, and the proportion of systems in which variation in $\gamma$ is critical for system stability. For each $S$, 1 million complex systems are randomly generated. Stability of each complex system is tested given variation in $\gamma$ by randomly sampling $\gamma \sim \mathcal{U}(0, 2)$. Stability given $Var(\gamma)$ is then compared to stability in an otherwise identical system in which $\gamma = E[\mathcal{U}(0, 2)]$ for all components. Light and dark grey bars show the number of stable systems in the absence and presence of $Var(\gamma)$, respectively. The black line shows the proportion of systems that are stable when $Var(\gamma) > 0$, but would be unstable if $Var(\gamma) = 0$.

@@

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

Given the larger space of Nature Communications, one new figure will be used in the same style as that of Figure 3; the new Figure 4 will show the same results, but for sultions found with the genetic algorithm.

Because Nature communcations is online, change all non-standard gamma results in figures to Red. The new genetic algorithm results will be in blue.

@@

Results
--------------------------------------------------------------------------------

@@

Discussion
--------------------------------------------------------------------------------

@@

Methods
--------------------------------------------------------------------------------

**Component response rate variation ($\mathbf{\gamma}$)**. In a synthesis of eco-evolutionary feedbacks on community stability, Patel et al. model a system that includes a vector of potentially changing species densities ($\mathbf{N}$) and a vector of potentially evolving traits ($\mathbf{x}$)[@Patel2018]. For any species $i$ or trait $j$, change in species density ($N_{i}$) or trait value ($x_{j}$) with time ($t$) is a function of the vectors $\mathbf{N}$ and $\mathbf{x}$,

$$\frac{dN_{i}}{dt} = N_{i}f_{i}(\mathbf{N}, \mathbf{x}),$$

$$\frac{dx_{j}}{dt} = \epsilon g_{j}(\mathbf{N}, \mathbf{x}).$$

In the above, $f_{i}$ and $g_{j}$ are functions that define the effects of all species densities and trait values on the density of a species $i$ and the value of trait $j$, respectively. Patel et al. were interested in stability when the evolution of traits was relatively slow or fast in comparison with the change in species densities[@Patel2018], and this is modulated in the above by the scalar $\epsilon$. The value of $\epsilon$ thereby determines the timescale separation between ecology and evolution, with high $\epsilon$ modelling relatively fast evolution and low $\epsilon$ modelling relative slow evolution[@Patel2018].

I use the same principle that Patel et al. use to modulate the relative rate of evolution to modulate rates of component responses for $S$ components. Following May[@May1972; @May1973], the value of a component $i$ at time $t$ ($v_{i}(t)$) is affected by the value of $j$ ($v_{j}(t)$) and $j$'s marginal effect on $i$ ($m_{ij}$), and by $i$'s response rate ($\gamma_{i}$),

$$\frac{dv_{i}(t)}{dt} = \gamma_{i} \sum_{j=1}^{S}m_{ij}v_{j}(t).$$

In matrix notation[@May1973],

$$\frac{d\mathbf{v}(t)}{dt} = \mathbf{\gamma} \mathbf{M}\mathbf{v}(t).$$

In the above, $\mathbf{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{\gamma} \mathbf{M}$ modulates the values of components and can be analysed using the techniques of May[@May1972; @May1973].

**Genetic algorithm**. Ideally, to investigate the potential of $Var(\gamma)$ for increasing the proportion of stable complex systems, the search space of all possible $\gamma$ vectors would be evaluated for each unique $M$. This is technically impossible because $\gamma_{i}$ can take any real value between 0-2, but even rounding $\gamma$ to reasonable values would result in a search space too large to practically explore. Under these conditions, genetic algorithms are highly useful tools for finding practical solutions by mimicking the process of biological evolution[@Hamblin2013]. In this case, the practical solution is finding vectors of $\gamma$ that decrease the most positive real eigenvalue of $M$. The genetic algorithm below achieves this by initialising a large population of 1000 different potential $\gamma$ vectors and allowing this population to evolve through a process of mutation, crossover (swaping $\gamma_{i}$ values between vectors), selection, and reproduction until either a $\gamma$ vector is found where all $\Re(\lambda) < 0$ or some "giving up" critiera is met (in the below, this "giving up"" criteria is met when 20 generations pass, or if the fitness increase from one generation to the next is below a certain criteria).

**System feasibility**. For complex systems in which individual system components ($S$) represent the density of some tangible quantity, it is important to consider the feasibility of the system. Feasibile equilibria assume that the values of all system components are positive at equilibrium[@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic[@Dougoud2018]. Consequently, the use of random matrices and traditional stability critiera for making inferences in theoretical analyses of species networks has recently been criticised[@Dougoud2018]. While the key results in the main text are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $M$. This analysis reveals that feasibility is not affected by $Var(\gamma)$, meaning that for pure interacting species networks, variation in component response time (i.e., species generation time) does not affect stability at biologically realistic species densities. Nevertheless, ecological interactions do not exist in isolation in empirical systems, but instead interact with evolutionary[@Patel2018], abiotic, or social-economic systems. The relevance of $\gamma$ for complex system stability presented in the main text should therefore not be ignored in the broader context of ecological communities.

Dougoud et al.[@Dougoud2018] define the following feasibility criteria for ecological systems characterised by $S$ interacting species with varying densities.

$$x^{*} = -\left(\theta I + (CS)^{-\delta}A\right)^{-1}r.$$

In the above, $x^{*}$ is the vector of species abundances at equilibrium (for feasibility, all values in $x^{*}$ must be positive). The matrix $I$ is the identity matrix (1s on the diagonal, 0s on the off-diagonal elements), and the value $\theta$ is strength of intraspecific competition (diagonal elements). As I have done elsewhere, diagonal values are set to $-1$, so $\theta = -1$. The variable $C$ is the inter-connectivity (i.e., 'connectance') of the community, which was set to $C = 1$ throughout the manuscript and supplemental information, except [where otherwise noted](#connectance). The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$ in the main text), which are held in $A$. In the main text, implicitly, $\delta = 0$ underlying strong interactions. Hence, the whole $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta I$) is added to $A$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's[@May1972] stability criterion[@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < -1$, and hence $(CS)^{-\delta} = 1$ for the randomly simulated systems in the main text and supplemental information). The above criteria is therefore reduced to the below; note that the parenthetical in both equations produces an $M$ matrix as used throughout the main text and supplemental information,

$$x^{*} = -\left(\theta I + A\right)^{-1}r.$$

To check the feasibility criteria, I therefore inverted $M = (\theta I + A)$ and multiplied elements by -1, then multiplied the resulting matrix by the vector of population growth rates $r$. Feasibility is satisfied if all of the elements of the resulting vector are positive.

The population growth rate for an individual species $i$ is sampled from a normal distribution of $r_{i} \sim \mathcal{N}(0, 0.4^{2})$.

@@

```{r, echo = FALSE}
# Temporary place for revised functions
plot_Fig_1 <- function(A0, A1){
    S_val       <- dim(A0)[1];
    A0_e        <- eigen(A0)$values;
    A0_r        <- Re(A0_e);
    A0_i        <- Im(A0_e);
    A1_e        <- eigen(A1)$values;
    A1_r        <- Re(A1_e);
    A1_i        <- Im(A1_e);
    A0_vm       <- A0;
    diag(A0_vm) <- NA;
    A0vec       <- as.vector(t(A0_vm));
    A0vec       <- A0vec[is.na(A0vec) == FALSE];
    A1_vm       <- A1;
    diag(A1_vm) <- NA;
    A1vec       <- as.vector(t(A1_vm));
    A1vec       <- A1vec[is.na(A1vec) == FALSE];
    fhalf       <- 1:(0.5*length(A1vec));
    shalf       <- (0.5*length(A1vec)+1):length(A1vec);
    par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
    plot(A0_r, A0_i, xlim = c(-3.7, 0.3), ylim = c(-2, 2), pch = 4, cex = 0.7,
         xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
    vl   <- seq(from = 0, to = 2*pi, by = 0.001);
    A0x0 <- sqrt(S_val) * sd(A0vec) * cos(vl) + mean(diag(A0));
    A0y0 <- sqrt(S_val) * sd(A0vec) * sin(vl);
    text(x = -3.5, y = 2.25, labels = "a", cex = 2);
    points(x = A0x0, y = A0y0, type = "l", lwd = 3, col = "dodgerblue4");
    points(A0_r, A0_i, pch = 4, cex = 0.7);
    plot(A1_r, A1_i, xlim = c(-3.7, 0.3), ylim = c(-2, 2), pch = 4, cex = 0.7,
         xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, 
         col = "dodgerblue4", yaxt = "n");
    vl <- seq(from = 0, to = 2*pi, by = 0.001);
    A0x1a <- sqrt(0.5*S_val) * sd(A1vec[fhalf]) * cos(vl) + 
                 mean(diag(A1)[1:(0.5*S_val)]);
    A0y1a <- sqrt(S_val) * sd(A1vec[fhalf]) * sin(vl);
    #points(x = A0x1a, y = A0y1a, type = "l", lwd = 3, col = "grey");
    A0x1b <- sqrt(0.5*S_val) * sd(A1vec[shalf]) * cos(vl) + 
        mean( diag(A1)[( (0.5*S_val) + 1 ):S_val] );
    A0y1b <- sqrt(0.5*S_val) * sd(A1vec[shalf]) * sin(vl);
    #points(x = A0x1b, y = A0y1b, type = "l", lwd = 3, col = "grey");
    points(A1_r[1:S_val], A1_i[1:S_val],pch = 4, cex = 0.7, col = "firebrick");   
    text(x = -3.5, y = 2.25, labels = "b", cex = 2);
    mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
    mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
}

plot_Fig_2 <- function(){
    A_comp <- NULL;
    A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
    A_mat  <- matrix(data = A_dat, nrow = 1000);
    C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 1);
    C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
    A_mat     <- A_mat * C_mat;
    gammas <- runif(n = 1000, min = 0, max = 2);
    mu_gam <- mean(gammas);
    diag(A_mat) <- -1;
    A1     <- gammas * A_mat;
    A0     <- mu_gam * A_mat;
    A0_e   <- eigen(A0)$values;
    A0_r   <- Re(A0_e);
    A0_i   <- Im(A0_e);
    A1_e   <- eigen(A1)$values;
    A1_r   <- Re(A1_e);
    A1_i   <- Im(A1_e);
    A0_vm       <- A0;
    diag(A0_vm) <- NA;
    A0vec       <- as.vector(A0_vm);
    A0vec       <- A0vec[is.na(A0vec) == FALSE];
    A1_vm       <- A1;
    diag(A1_vm) <- NA;
    A1vec       <- as.vector(A1_vm);
    A1vec       <- A1vec[is.na(A1vec) == FALSE];
    par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
    plot(A0_r, A0_i, xlim = c(-16.5, 15.5), ylim = c(-16.5,15.5), pch = 4, 
         cex = 0.7, xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, 
         asp = 1, col = "dodgerblue4");
    vl <- seq(from = 0, to = 2*pi, by = 0.001);
    x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
    y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
    x1 <- sqrt(1000) * sd(A1vec) * cos(vl) + mean(diag(A1));
    y1 <- sqrt(1000) * sd(A1vec) * sin(vl);
    text(x = -15.5, y = 19, labels = "a", cex = 2);
    points(x = x0, y = y0, type = "l", lwd = 3, col = "dodgerblue4");
    points(x = x1, y = y1, type = "l", col = "red", lwd = 3, lty = "dashed");
    plot(A1_r, A1_i, xlim = c(-16.5, 15.5), ylim = c(-16.5,15.5), pch = 4, 
         cex = 0.7, xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, 
         asp = 1, col = "firebrick", yaxt = "n");
    text(x = -15.5, y = 19, labels = "b", cex = 2);
    points(x = x1, y = y1, type = "l", col = "firebrick", lwd = 3);
    points(x = x0, y = y0, type = "l", lwd = 3, lty = "dashed");
    mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
    mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
}

plot_stables <- function(dat, S_s = 32){
    Ns          <- 1:S_s;
    par(oma = c(6, 6, 1, 6), mar = c(0.5, 0.5, 0.5, 0.5));
    #=================================
    bar_dat                      <- t(cbind(dat[Ns,3], dat[Ns,5]));
    log_bar_dat                  <- log(bar_dat);
    log_bar_dat[log_bar_dat < 0] <- 0; 
    barplot(log_bar_dat, beside = TRUE, col = c("dodgerblue4", "firebrick"),
            names.arg = dat[Ns,1], ylim = c(0, 16), xlab = "",
            ylab = "Ln number of stable communities", cex.lab = 1, 
            cex.axis = 1.25, xlim = c(1, 94), cex.names = 1, yaxt = "n");
    axis(side = 2, at = c(0, 2, 4, 6, 8, 10, 12, 14), cex.axis = 1.5);
    box(lwd = 2);
    par(new = TRUE);
    y1     <- dat[1:S_s,6] / (dat[1:S_s,5]);
    x1     <- seq(from = 2.132, to = 15.1112, length = S_s);
    plot(x = x1, y = y1, xaxt = "n", yaxt = "n", lwd = 2, ylim = c(0, 1.1),
         xlab = "", ylab = "", type = "b", xlim = c(2, 15), pch = 20, 
         cex = 1, col = "black", yaxs="i");
    points(x = x1, y = y1, lwd = 2, type = "l", col = "black");
    axis(side = 4, at = c(0, 0.2, 0.4, 0.6, 0.8, 1.0), cex.axis = 1.5);
    legend("topleft", c(expression(paste(gamma," = 1")), 
                        expression(paste("Var(",gamma,")"))), 
           pch=15, col=c("dodgerblue4","firebrick"), cex = 1.5, horiz = TRUE);
    #=================================
    mtext(side = 1, text = "System size (S)", cex = 2, outer = TRUE, 
          line = 3.0);
    mtext(side = 2, text = "Ln number of stable systems", cex = 2, 
          outer = TRUE, line = 3.5);
    mtext(side = 4, text = expression(
        paste("Pr. of systems stable due to Var(",gamma,")")),
        cex = 2, outer = TRUE, line = 3.5);
}

plot_stable_4 <- function(dat, S_s = 39){
    Ns          <- 1:S_s;
    par(oma = c(6, 6, 1, 6), mar = c(0.5, 0.5, 0.5, 0.5));
    #=================================
    bar_dat                      <- t(cbind(dat[Ns,3], dat[Ns,5]));
    log_bar_dat                  <- log(bar_dat);
    log_bar_dat[log_bar_dat < 0] <- 0; 
    barplot(log_bar_dat, beside = TRUE, col = c("dodgerblue4", "firebrick"),
            names.arg = dat[Ns,1], ylim = c(0, 14), xlab = "",
            ylab = "Ln number of stable communities", cex.lab = 1, 
            cex.axis = 1.25, xlim = c(1, 116), cex.names = 1, yaxt = "n");
    axis(side = 2, at = c(0, 2, 4, 6, 8, 10, 12), cex.axis = 1.5);
    box(lwd = 2);
    par(new = TRUE);
    y1     <- dat[1:S_s,6] / (dat[1:S_s,5]);
    x1     <- seq(from = 2.132, to = 15.1112, length = S_s);
    plot(x = x1, y = y1, xaxt = "n", yaxt = "n", lwd = 2, ylim = c(0, 1.1),
         xlab = "", ylab = "", type = "b", xlim = c(2, 15), pch = 20, 
         cex = 1, col = "black", yaxs="i");
    points(x = x1, y = y1, lwd = 2, type = "l", col = "black");
    axis(side = 4, at = c(0, 0.2, 0.4, 0.6, 0.8, 1.0), cex.axis = 1.5);
    legend("topleft", c(expression(paste(gamma," = 1")), 
                        expression(paste("Var(",gamma,")"))), 
           pch=15, col=c("dodgerblue4","firebrick"), cex = 1.5, horiz = TRUE);
    #=================================
    mtext(side = 1, text = "System size (S)", cex = 2, outer = TRUE, 
          line = 3.0);
    mtext(side = 2, text = "Ln number of stable systems", cex = 2, 
          outer = TRUE, line = 3.5);
    mtext(side = 4, text = expression(
        paste("Pr. of systems stable due to Var(",gamma,")")),
        cex = 2, outer = TRUE, line = 3.5);
}
```

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the blue circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle centred at $\gamma = -1.95$ and one small circle centred at $\gamma = -0.05$. In a, the maximum real eigenvalue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` was stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-3.7, 0.3), ylim = c(-2, 2), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, 
     col = "dodgerblue4");
vl <- seq(from = 0, to = 2*pi, by = 0.001);
A0x0 <- sqrt(200) * sd(A0vec) * cos(vl) + mean(diag(A0));
A0y0 <- sqrt(200) * sd(A0vec) * sin(vl);
text(x = -3.5, y = 2.25, labels = "a", cex = 2);
points(x = A0x0, y = A0y0, type = "l", lwd = 3, col = "blue");
points(A0_r, A0_i, pch = 4, cex = 0.7, col = "dodgerblue4");

plot(A1_r, A1_i, xlim = c(-3.7, 0.3), ylim = c(-2, 2), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, 
     col = "firebrick", yaxt = "n");

vl <- seq(from = 0, to = 2*pi, by = 0.001);
A0x1a <- sqrt(100) * sd(A1vec[fhalf]) * cos(vl) + mean(diag(A1)[1:100]);
A0y1a <- sqrt(100) * sd(A1vec[fhalf]) * sin(vl);
points(x = A0x1a, y = A0y1a, type = "l", lwd = 3, col = "red");
A0x1b <- sqrt(100) * sd(A1vec[shalf]) * cos(vl) + mean(diag(A1)[101:200]);
A0y1b <- sqrt(100) * sd(A1vec[shalf]) * sin(vl);
points(x = A0x1b, y = A0y1b, type = "l", lwd = 3, col = "red");

points(A1_r[1:100], A1_i[1:100],pch = 4, cex = 0.7, col = "firebrick");

@@

**Figure 2: Distributions of eigenvalues before (a) and after (b) introducing variation in component response rate ($\boldsymbol{\gamma}$) in complex systems.** Each panel show the same system where $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. **a.** Eigenvalues plotted in the absence of $Var(\gamma)$ where $E[\gamma] = 1$, versus **b.** eigenvalues plotted given $\gamma \sim \mathcal{U}(0, 2)$, which increases the variance of interaction strengths ($\sigma^{2}$) but also creates a cluster of eigenvalues toward the distribution's centre (-1, 0). Blue elipses in both panels show the circle centred on the distribution in panel a. Proportions of $\Re(\lambda) < 0$ are `r sum(A0_r < 0) / length(A0_r)` and `r sum(A1_r < 0) / length(A1_r)` for a and b, respectively.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-6, 2.25), ylim = c(-5,4.25), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, 
     col = "dodgerblue4");
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
text(x = -15.5, y = 19, labels = "a", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3, col = "blue");
points(A0_r, A0_i, col = "dodgerblue4",pch = 4, cex = 0.7);
plot(A1_r, A1_i, xlim = c(-6, 2.25), ylim = c(-5,4.25), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, 
     col = "firebrick", yaxt = "n");
text(x = -15.5, y = 19, labels = "b", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3, lty = "solid", col = "blue");
points(A1_r, A1_i, col = "firebrick", pch = 4, cex = 0.7);
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

@@

**Figure 3: Stability of large complex systems with and without variation in component response rate ($\boldsymbol{\gamma}$).** The $\ln$ number of systems that are stable across different system sizes ($S$) given $C = 1$, and the proportion of systems in which variation in $\gamma$ is critical for system stability. For each $S$, 1 million complex systems are randomly generated. Stability of each complex system is tested given variation in $\gamma$ by randomly sampling $\gamma \sim \mathcal{U}(0, 2)$. Stability given $Var(\gamma)$ is then compared to stability in an otherwise identical system in which $\gamma = E[\mathcal{U}(0, 2)]$ for all components. Blue and red bars show the number of stable systems in the absence and presence of $Var(\gamma)$, respectively. The black line shows the proportion of systems that are stable when $Var(\gamma) > 0$, but would be unstable if $Var(\gamma) = 0$.

@@

**Figure 4: Stability of large complex systems given $\boldsymbol{\gamma = 1}$ versus targetted $\boldsymbol{Var(\gamma)}$.** The $\ln$ number of systems that are stable across different system sizes ($S$, max $S=40$) for $C = 1$, and the proportion of systems wherein a targetted search of $\gamma$ values successfully resulted in system stability. For each $S$, 100000 complex systems are randomly generated. Stability of each complex system is tested given variation in $\gamma$ using a genetic algorithm to maximise the effect of $\gamma$ values on increasing stability, as compared to stability in an otherwise identical system in which $\gamma$ is the same for all components. Blue bars show the number of stable systems in the absence of component response rate variation, while red bars show the number of stable systems that can be generated if component response rate is varied to maximise system stability. The black line shows the proportion of systems that are stable when component response rate is targetted to increase stability, but would not be stable if $Var(\gamma) = 0$.

```{r, eval = TRUE, echo = FALSE, fig.width = 9, fig.height = 7}
dat <- read.csv(file = "sim_results/evolved/evo_results.csv");
dat <- dat[,-1];
plot_stable_4(dat);
```
